---
title: "STATS_506_PS2"
author: "Zixun Zhang"
format: typst
editor: visual
---

**Github repository link:** <https://github.com/zixunzhang99/stats506>

## **Problem 1: Random Walk with Momentum**

### **a. Simulation function:**

```{r}
#' Simulate a two-dimensional random walk with momentum
#'
#' Simulates a random walk starting at (0, 0). The first direction is
#' chosen randomly. For each subsequent step, the previous direction is
#' preserved with probability p; otherwise, one of the four directions
#' is chosen randomly.
#'
#' @param n A positive integer representing the number of steps.
#' @param p A numeric value between 0 and 1 representing the probability
#'   of preserving the previous direction.
#'
#' @return A numeric vector of length two containing the final x and y
#'   coordinates of the particle.
momentum_walk <- function(n, p) {
  if (!is.numeric(n) || length(n) != 1 || n < 1 || n %% 1 != 0) {
    stop("n must be a positive integer")
  }
  
  if (!is.numeric(p) || length(p) != 1 || p < 0 || p > 1) {
    stop("p must be a number between 0 and 1")
  }
  
  directions <- matrix(
    c(1, 0,
      -1, 0,
      0, 1,
      0, -1),
    nrow = 4,
    byrow = TRUE
  )
  
  position <- c(0, 0)
  
  current_dir <- sample(1:4, 1)
  position <- position + directions[current_dir, ]
  
  if (n > 1) {
    for (i in 2:n) {
      
      if (runif(1) > p) {
        current_dir <- sample(1:4, 1)
      }

      position <- position + directions[current_dir, ]
    }
  }
  
  return(position)
}
```

```{r}
momentum_walk(10, 0.6)
momentum_walk(10, 0.6)
```

### **b. Implement the simulation**

#### i. Version 1: write the function `momentum_walk1(n, p)` using a loop ( a `for` loop or a `while` loop).

```{r}
#' Simulate a two-dimensional random walk with momentum
#'
#' Simulates a random walk starting at (0, 0). The first direction is
#' chosen randomly. For each subsequent step, the previous direction is
#' preserved with probability p; otherwise, one of the four directions
#' is chosen randomly.
#'
#' @param n A positive integer representing the number of steps.
#' @param p A numeric value between 0 and 1 representing the probability
#'   of preserving the previous direction.
#'
#' @return A numeric vector of length two containing the final x and y
#'   coordinates of the particle.
momentum_walk1 <- function(n, p) {
  if (!is.numeric(n) || length(n) != 1 || n < 1 || n %% 1 != 0) {
    stop("n must be a positive integer")
  }
  
  if (!is.numeric(p) || length(p) != 1 || p < 0 || p > 1) {
    stop("p must be a number between 0 and 1")
  }
  
  directions <- matrix(
    c(1, 0,
      -1, 0,
      0, 1,
      0, -1),
    nrow = 4,
    byrow = TRUE
  )
  
  position <- c(0, 0)
  
  current_dir <- sample(1:4, 1)
  position <- position + directions[current_dir, ]
  
  if (n > 1) {
    for (i in 2:n) {
      
      if (runif(1) > p) {
        current_dir <- sample(1:4, 1)
      }
      
      position <- position + directions[current_dir, ]
    }
  }
  
  return(position)
}
```

For `momentum_walk1()`, a `for` loop is used to simulate the walk one step at a time. The first direction is selected randomly. For each subsequent step, the function either preserves the previous direction with probability (p), or randomly selects one of the four directions with probability (1-p). The particle's position is updated after each step.

#### ii. Version 2: write the function `momentum_walk2(n, p)` using one of the apply functions (choose an appropriate member of the apply family).

```{r}
#' Simulate a random walk with momentum using an apply function
#'
#' Simulates a two-dimensional random walk starting at the origin.
#' The first direction is chosen randomly. For each subsequent step,
#' the particle preserves its previous direction with probability `p`;
#' otherwise, a new direction is chosen randomly from the four possible
#' directions.
#'
#' @param n A positive integer specifying the number of steps.
#' @param p A numeric value between 0 and 1 specifying the probability
#'   of preserving the previous direction.
#'
#' @return A numeric vector of length 2 containing the final x and y
#'   coordinates of the particle.
momentum_walk2 <- function(n, p) {
  if (!is.numeric(n) || length(n) != 1 || n < 1 || n %% 1 != 0) {
    stop("n must be a positive integer")
  }
  
  if (!is.numeric(p) || length(p) != 1 || p < 0 || p > 1) {
    stop("p must be a number between 0 and 1")
  }
  
  directions <- matrix(
    c(1, 0,
      -1, 0,
      0, 1,
      0, -1),
    nrow = 4,
    byrow = TRUE
  )
  
  current_dir <- sample(1:4, 1)
  
  position <- directions[current_dir, ]
  
  if (n == 1) {
    return(position)
  }
  
  steps <- sapply(2:n, function(i) {
    
    if (runif(1) >= p) {
      current_dir <<- sample(1:4, 1)
    }
    
    directions[current_dir, ]
  })
  
  position <- position + rowSums(steps)
  
  return(position)
}
```

For `momentum_walk2()`, `sapply()` is used to generate the subsequent steps instead of an explicit loop. The function similarly keeps track of the previous direction and either preserves it or randomly selects a new direction at each step. The resulting movements are then summed to obtain the final position.

```{r}
momentum_walk1(10, 0.6)
momentum_walk2(10, 0.6)
```

```{r}
momentum_walk1(1000, 0.6)
momentum_walk2(1000, 0.6)
```

### **c. Reproducibility**

```{r}
set.seed(506)

momentum_walk1(10, 0.6)
momentum_walk1(1000, 0.6)
```

I used `set.seed()` to set the initial state of R's random-number generator. Using the same seed before running the simulation causes R to generate the same sequence of random numbers, so the simulation produces the same result and is reproducible.

### **d. Does momentum matter?**

#### i. For each value of p, simulate a walk of 1000 steps.

```{r}
set.seed(506)

walk_0 <- momentum_walk1(1000, 0)
walk_05 <- momentum_walk1(1000, 0.5)
walk_09 <- momentum_walk1(1000, 0.9)

dist_0 <- sqrt(sum(walk_0^2))
dist_05 <- sqrt(sum(walk_05^2))
dist_09 <- sqrt(sum(walk_09^2))

data.frame(
  p = c(0, 0.5, 0.9),
  distance = c(dist_0, dist_05, dist_09)
)
```

#### ii. Does increasing momentum appear to increase or decrease the final distance?

Increasing momentum appears to increase the final distance.

#### iii. Explain your reasoning

With larger momentum, the particle is more likely to preserve its previous direction, resulting in longer runs in the same direction.

### **e. Monte Carlo simulation**

#### i. - iii.

```{r}
set.seed(506)

p_values <- c(0, 0.25, 0.5, 0.75, 0.9)

mean_distances <- sapply(p_values, function(p) {
  
  distances <- replicate(1000, {
    
    final_pos <- momentum_walk1(1000, p)
    
    sqrt(sum(final_pos^2))
  })
  
  mean(distances)
})

results <- data.frame(
  p = p_values,
  mean_distance = mean_distances
)

results
```

#### iv. Answer: How does the mean final distance change as p increases?

The mean final distance increases as p increases.

### **f. Compare computational performance**

```{r}
library(microbenchmark)
```

```{r}
benchmark_1000 <- microbenchmark(
  loop = momentum_walk1(1000, 0.6),
  apply = momentum_walk2(1000, 0.6)
)

benchmark_1000
```

```{r}
benchmark_100000 <- microbenchmark(
  loop = momentum_walk1(100000, 0.6),
  apply = momentum_walk2(100000, 0.6)
)

benchmark_100000
```

#### i. - iii.

The `for` loop implementation is faster. Yes, the difference becomes larger for the larger input. This may occur because the apply implementation introduces additional function-call and result-handling overhead. Since each step of the random walk depends on the previous step, the apply implementation does not provide a substantial vectorization advantage over the loop.

## **Problem 2: Monte Carlo Simulation of Campus Bus Arrivals**

```{r}
set.seed(506)

n_days <- 10000

morning <- matrix(
  rpois(n_days * 3, lambda = 4),
  nrow = n_days
)

peak_morning <- rnorm(
  n_days,
  mean = 45,
  sd = sqrt(16)
)

midday <- matrix(
  rpois(n_days * 6, lambda = 15),
  nrow = n_days
)

peak_afternoon <- rnorm(
  n_days,
  mean = 55,
  sd = sqrt(25)
)

evening <- matrix(
  rpois(n_days * 6, lambda = 9),
  nrow = n_days
)
```

```{r}
daily_total <- rowSums(morning) +
  peak_morning +
  rowSums(midday) +
  peak_afternoon +
  rowSums(evening)

mean(daily_total)
```

The expected total number of passengers who board the bus per day is about 256.

## **Problem 3: Linear Regression**

```{r}
youtube <- read.csv(
  "https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-03-02/youtube.csv"
)
```

### **a. Data preparation and de-identification**

```{r}
remove_cols <- c(
  "brand",
  "superbowl_ads_dot_com_url",
  "youtube_url",
  "id",
  "kind",
  "etag",
  "published_at",
  "title",
  "description",
  "thumbnail",
  "channel_title"
)

youtube_clean <- youtube[, !(names(youtube) %in% remove_cols)]

dim(youtube_clean)
```

### **b. Exploring potential outcomes**

```{r}
library(ggplot2)

ggplot(youtube_clean, aes(x = view_count)) +
  geom_histogram() +
  labs(
    x = "View Count",
    y = "Frequency"
  )

ggplot(youtube_clean, aes(x = like_count)) +
  geom_histogram() +
  labs(
    x = "Like Count",
    y = "Frequency"
  )

ggplot(youtube_clean, aes(x = dislike_count)) +
  geom_histogram() +
  labs(
    x = "Dislike Count",
    y = "Frequency"
  )

ggplot(youtube_clean, aes(x = favorite_count)) +
  geom_histogram() +
  labs(
    x = "Favorite Count",
    y = "Frequency"
  )

ggplot(youtube_clean, aes(x = comment_count)) +
  geom_histogram() +
  labs(
    x = "Comment Count",
    y = "Frequency"
  )
```

```{r}
summary(youtube_clean$view_count)
summary(youtube_clean$like_count)
summary(youtube_clean$dislike_count)
summary(youtube_clean$favorite_count)
summary(youtube_clean$comment_count)
```

```{r}
youtube_clean$log_view_count <- log(youtube_clean$view_count)

youtube_clean$sqrt_like_count <- sqrt(youtube_clean$like_count)

youtube_clean$sqrt_dislike_count <- sqrt(youtube_clean$dislike_count)

youtube_clean$sqrt_comment_count <- sqrt(youtube_clean$comment_count)
```

```{r}
ggplot(youtube_clean, aes(x = log_view_count)) +
  geom_histogram() +
  labs(x = "Log View Count", y = "Frequency")

ggplot(youtube_clean, aes(x = sqrt_like_count)) +
  geom_histogram() +
  labs(x = "Square Root Like Count", y = "Frequency")

ggplot(youtube_clean, aes(x = sqrt_dislike_count)) +
  geom_histogram() +
  labs(x = "Square Root Dislike Count", y = "Frequency")

ggplot(youtube_clean, aes(x = sqrt_comment_count)) +
  geom_histogram() +
  labs(x = "Square Root Comment Count", y = "Frequency")
```

The distribution of `view_count` is extremely right-skewed, with a small number of observations having very large view counts. Therefore, I applied a log transformation. The transformed distribution is substantially more symmetric and is more appropriate for use as an outcome in linear regression.

The distribution of `like_count`, `dislike_count`, and `comment_count` are also heavily right-skewed and contains zeros. I applied a square-root transformation to reduce the skewness while allowing zero values to remain in the data.

`favorite_count` is not appropriate as an outcome for linear regression because all non-missing observations have a value of zero. Therefore, the variable has no variation to model.

### **c. Fitting linear regression models**

```{r}
characteristics <- c(
  "funny",
  "show_product_quickly",
  "patriotic",
  "celebrity",
  "danger",
  "animals",
  "use_sex"
)

fit_model <- function(outcome, characteristic) {
  
  formula <- as.formula(
    paste(outcome, "~", characteristic, "+ year")
  )
  
  model <- lm(formula, data = youtube_clean)
  
  results <- summary(model)$coefficients
  
  c(
    estimate = results[2, "Estimate"],
    p_value = results[2, "Pr(>|t|)"]
  )
}
```

```{r}
view_results <- t(sapply(
  characteristics,
  function(x) fit_model("log_view_count", x)
))

view_results
```

Since all the p-values are larger than 0.05, none of the characteristics is significant. The commercial characteristic has potentially the largest positive estimated association with log(view_counts) is `danger`.

```{r}
like_results <- t(sapply(
  characteristics,
  function(x) fit_model("sqrt_like_count", x)
))

like_results
```

Since all the p-values are larger than 0.05, none of the characteristics is significant.

```{r}
dislike_results <- t(sapply(
  characteristics,
  function(x) fit_model("sqrt_dislike_count", x)
))

dislike_results
```

Since all the p-values are larger than 0.05, none of the characteristics is significant.

```{r}
comment_results <- t(sapply(
  characteristics,
  function(x) fit_model("sqrt_comment_count", x)
))

comment_results
```

The p-value for `patriotic` is less than 0.05, thus it has a statistically significant association with sqrt(comment_count). Since it is about 4.534, it has the largest and a positive association.

### **d. Estimating regression coefficients using matrix algebra**

```{r}
dat <- youtube_clean[
  complete.cases(youtube_clean[, c("log_view_count", "danger", "year")]),
]
```

```{r}
X <- model.matrix(
  ~ danger + year,
  data = dat
)

y <- dat$log_view_count

beta_hat <- solve(t(X) %*% X) %*% t(X) %*% y

beta_hat
```

```{r}
model <- lm(
  log_view_count ~ danger + year,
  data = dat
)

coef(model)
```
