---
title: "STATS_506_PS1"
author: "Zixun Zhang"
format: typst
editor: visual
---

## **Problem 1 - Exploratory Analysis of Soybean Cultivars**

#### a. Import the data into a `data.frame` in R. Use the dataset documentation to assign appropriate variable names.

```{r}
soybean <- read.csv("data.csv")

names(soybean) <- c(
  "season",
  "cultivar",
  "rep",
  "height",
  "first_pod",
  "stems",
  "legumes",
  "grains",
  "grains_pod",
  "seed_weight",
  "yield"
)

head(soybean)
```

#### b. Using the information provided in the dataset documentation: Verify that there are 320 observations in the dataset; 40 distinct cultivars; and all the observations were collected from two seasons. Produce a table showing the number of observations from each season.

```{r}
obs_num <- nrow(soybean)

cul_num <- length(unique(soybean$cultivar))

seasons <- unique(soybean$season)

obs_num
cul_num
seasons
table(soybean$season)
```

#### c. Use the data to answer the following questions:

1.  What is the correlation between plant height (PH) and grain yield (GY)?

2.  Calculate the correlation between PH and GY separately for each season. Which season has the higher correlation?

3.  What is the grain yield of the soybean observation with the greatest plant height?

4.  What percentage of observations have a thousand seed weight (MHG) greater than the overall average thousand seed weight?

5.  Which cultivar has the highest average grain yield?

```{r}
# 1
print("Correlation between plant height and grain yield:")
cor(soybean$height, soybean$yield)

# 2
print("Correlation between plant height and grain yield for season 1:")
cor(soybean$height[soybean$season == 1],
    soybean$yield[soybean$season == 1])

print("Correlation between plant height and grain yield for season 2:")
cor(soybean$height[soybean$season == 2],
    soybean$yield[soybean$season == 2])

# 3
print("Grain yield of the soybean observation with the greatest plant height")
soybean$yield[which.max(soybean$height)]

# 4
print("Percentage of observations have a thousand seed weight (MHG) greater than the overall average")
mean(soybean$seed_weight > mean(soybean$seed_weight)) * 100

# 5
print("Cultivar with the highest average grain yield:")
avg_yield <- tapply(soybean$yield, soybean$cultivar, mean)
names(avg_yield)[which.max(avg_yield)]
```

1.  The correlation between plant height and grain yield is about 0.1233.
2.  The correlation between plant height and grain yield for season 1 is about 0.2309, and about -0.0227 for season 2.
3.  The grain yield of the soybean observation with the greatest plant height is 3304.678 kg/ha.
4.  About 47.19% observations have a thousand seed weight greater than the overall average thousand seed weight.
5.  "98R30 CE" has the highest average grain yield.

#### d. Create a table showing the average value of each quantitative variable:

1.  One row containing overall averages.

2.  One row for each season containing season-specific averages.

```{r}
quant <- soybean[, c(
  "height", "first_pod", "stems", "legumes",
  "grains", "grains_pod", "seed_weight", "yield"
)]

overall <- sapply(quant, mean)

season1 <- sapply(quant[soybean$season == 1, ], mean)
season2 <- sapply(quant[soybean$season == 2, ], mean)

avg_table <- rbind(
  Overall = overall,
  Season1 = season1,
  Season2 = season2
)

avg_table
```

#### e. Carry out a series of t-tests to examine whether the grain yield (GY) differs between the two seasons.

1.  State the null and the alternative hypotheses.

2.  Present the R output and interpret the results.

3.  The null hypothesis would be the grain yields are the same between the two seasons. The alternative hypothesis is that the grain yields are different between the two seasons.

```{r}
season1_yield <- soybean$yield[soybean$season == 1]
season2_yield <- soybean$yield[soybean$season == 2]

t.test(season1_yield, season2_yield)
```

2.  Assuming the grain yields are the same between the two seasons, the probability of observing a t statistics of 0.352 or more further away on either direction is 0.7255. Using a significance level of 5%, the data do not provide enough evidence to conclude that the grain yields are different across the two seasons.

## **Problem 2 - Adult Income Dataset**

#### a. Import the data into a `data.frame` in R. As with the previous data, you may download the data outside of your submission, but importation should take place inside the problem set submission.

```{r}
adult <- read.csv("adult.data", header = FALSE)
head(adult)
```

#### b. Clean and simplify the variable names.

```{r}
names(adult) <- c(
  "age",
  "work_class",
  "final_weight",
  "edu",
  "yr_of_edu",
  "marital",
  "occupation",
  "relationship",
  "race",
  "sex",
  "gain",
  "loss",
  "hr_per_week",
  "native_country",
  "income"
)

head(adult)
```

#### c. Examine missing values. Report the number of missing observations for each variable.

```{r}
sapply(adult, function(x){
  sum(x == " ?")
})
```

#### d. Restrict the sample to individuals working at least 10 hours per week. Confirm the sample size before and after filtering.

```{r}
print("Sample size before filtering:")
nrow(adult)

print("Sample size after filtering:")
working_adult <- adult[adult$hr_per_week >= 10, ]
nrow(working_adult)
```

#### e. Remove observations with missing values for occupation, workclass, or native country. Explain why.

```{r}
missing_data <- " ?"

working_adult <- working_adult[
  working_adult$occupation != missing_data & 
  working_adult$work_class != missing_data &
  working_adult$native_country != missing_data,
]
```

We are removing observations with missing values because they do not provide complete information for the subsequent analysis.

#### f. Eliminate any observations with: age less than 18, age greater than 90, education level identified as Preschool.

```{r}
working_adult <- working_adult[
  working_adult$age >= 18 & 
  working_adult$age <= 90 &
  working_adult$edu != "Preschool",
]
```

#### g. Report your final sample size.

```{r}
print("The final sample size is:")
nrow(working_adult)
```

The final sample size is 29568.

## **Problem 3 - Happy Numbers**

#### a. Write a function `nextHappy()` that given a positive integer, computes the next value in the happy-number sequence. Be sure to provide a reasonable error on an invalid input. Be sure to document your function

```{r}
#' Calculate the next happy number.
#'
#' Computes the next value in a happy-number sequence by summing
#' the squares of the digits of a positive integer.
#' 
#' @param n  A positive integer.
#'
#' @returns The sum of the squared digits of `n`.
nextHappy <- function(n) {
  if (!is.numeric(n) || length(n) != 1) {
    stop("n must be a single numeric value")
  }
  if (n <= 0 || n %% 1 != 0) {
    stop("n must be a positive integer")
  }
  
  total <- 0
  
  while (n > 0) {
    digit <- n %% 10
    total <- total + digit^2
    n <- n %/% 10
  }
  
  return(total)
}
```

```{r}
nextHappy(19)
nextHappy(82)
```

#### b. Create a function `happySequence()` that returns the sequence beginning at a given number. Use your `nextHappy()` function to perform the calculation. Be sure to provide a reasonable error on an invalid input. Be sure to document your functio

```{r}
#' Identify if a number is a happy number.
#'
#' Determines whether a positive integer is a happy number. If the number is
#' happy, the happy-number sequence leading to 1 is also returned.
#'
#' @param n A positive number
#'
#' @returns A list containing `isHappy`, a logical value indicating whether
#'   `n` is a happy number. If `n` is happy, the list also contains `sequence`,
#'   a numeric vector giving the sequence from `n` to 1.
happySequence <- function(n) {
  
  if (!is.numeric(n) || length(n) != 1) {
    stop("n must be a single numeric value")
  }
  
  if (n <= 0 || n %% 1 != 0) {
    stop("n must be a positive integer")
  }
  
  prev <- c()
  current <- n

  
  while (!(current %in% prev)) {
    
    prev <- c(prev, current)
    
    if (current == 1) {
      return(return(list(
        sequence = prev,
        isHappy = TRUE
      )))
    }
    current <- nextHappy(current)
  }
  
  return(list(
    isHappy = FALSE
  ))
}
```

```{r}
happySequence(19)
happySequence(4)
```

#### c. Among integers 1 through 1000:

a.  How many are happy?

b.  What percentage are happy?

c.  What starting value produces the longest happy sequence before terminating?

```{r}
nums <- 1:1000

happy_seq <- sapply(nums, function(x) {
  happySequence(x)$isHappy
})

happy_count <- sum(happy_seq)
happy_count

happy_percentage <- happy_count / length(nums) * 100
happy_percentage

happy_seq_len <- sapply(nums, function(x) {
  length(happySequence(x)$sequence)
})
longest_seq_val <- nums[which.max(happy_seq_len)]
longest_seq_val
```

a\. 143 numbers are happy numbers

b\. 14.3% numbers are happy numbers

c\. 356 produces the longest happy sequence before terminating.

**Github repository link:** <https://github.com/zixunzhang99/stats506>
